---
title: 'Lab 1:'
author: "Sachiko Lamen"
date: "1/6/2022"
output: html_document
---

```{r setup, include=TRUE, warning = FALSE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(tidyverse)
library(here)
library(sf)
library(tmap)
```

cmd-shift-enter: shortcut for running the current code chunk

### Read in the data
cmd-option-i: shortcut for creating a code chunk

```{r}
sf_trees <- read_csv(here("data", "sf_trees", "sf_trees.csv"),
                     show_col_types = FALSE) # hides information about column types that usually pops up under code chunk 

# In console:
# view(sf_trees) to view dataset
# names(sf_trees) will give names of each column
# summary(sf_trees) gives summary of min, max, mean, etc of each column <- quick rundown of values
```

### Part 1: Wrangling and ggplot review

**Example 1:** Find counts of observations by 'legal_status' & wrangle a bit.

```{r}
### Method 1: group_by %>% summarize()
sf_trees %>%
  group_by(legal_status) %>%
  summarize(tree_count = n()) # n() counts how many rows there are in the data frame

### Method 2: different way, plus a few new funcitons
top_five_status <- sf_trees %>%
  count(legal_status) %>%
  drop_na(legal_status) %>% # will drop any row with an 'NA' in it in the legal_status column 
  rename(tree_count = n) %>% 
  relocate(tree_count) %>% # relocates tree_count to the front of the columns
  slice_max(tree_count, n = 5) %>% # takes only maximum up to 5
  arrange(desc(tree_count)) # arrange by tree_count in descending order

```

### Make a graph of the top 5 from above:

```{r}
ggplot(data = top_five_status, aes(x = fct_reorder(legal_status, tree_count), y = tree_count)) +
         geom_col(fill = 'darkgreen') +
  labs(x = 'Legal Status',
       y = 'Tree Count') +
  coord_flip() +
  theme_minimal()
       
# use 'fct_reorder() to reoder 'legal_status' by 'tree_count' and order by factor (from smallest to largest) ** Note: you can use '-' infront of 'tree_count' to switch order to go from highest to lowest 
```

**Example 2** Only going to keep observations where legal status is "Permitted Site" and caretaker is "MTA", and store as `permitted_data_df`

shift-cmd-c to commen/uncomment quickly
```{r}
# sf_trees$legal_status %>% unique() # takes legal status column and gets rid of duplicates to see what the unique values are
# unique(sf_trees$caretaker)

permitted_data_df <- sf_trees %>%
  filter(legal_status == 'Permitted Site', caretaker == 'MTA') 

# Note: '|' between 'Permitted Site' and 'caretaker' means 'or', so if legal status is 'Permitted Site' OR caretaker is 'MTA', or both then it should appear in the data table. EX) filter(legal_status == 'Permitted Site' | caretaker == 'MTA') 

#Note: if you want to filter for multiple legal statuses use %in% and `c()` EX) filter(legal_status %in% c('Permitted Site', 'Private'), caretaker == 'MTA')
```

**Example 3:** Only keep Blackwood Acacia trees, and then only keep columns `legal status`, `date`, `latitude`, `longitude`, and store as `blackwood_acacia_df`

```{r}
blackwood_acacia_df <- sf_trees %>%
  filter(str_detect(species, 'Blackwood Acacia')) %>% # in this column, keep any rows where you find 'Blackwood Acacia'
  select(legal_status, date, lat = latitude, lon = longitude)

### Make a little graph of locations

ggplot(data = blackwood_acacia_df, aes(x = lon, y = lat)) + 
  geom_point(color = 'darkgreen')
```

**Example 4** Use `tidyr::separate()` to separate one column into multiple columns

```{r}
sf_trees_sep <- sf_trees %>%
  separate(species, into = c("species_scientific", "species_common"), sep = "::") # separate 'species' column into two columns at the '::'
```

**Example 5** use `tidyr::unite()` to combine columns

```{r}
ex_5 <- sf_trees %>%
  unite('id_status', tree_id, legal_status, sep = '_CHECK_THIS_OUT')

# Note: also drops the old columns and only retains the new column, 'id_status'
```











